Scalable Clustering of Genotype Information using MapReduce

### 2014-05-01 doc/publications/incob2014/paper.tex
intro

As a result of these big data challenges, MapReduce approaches are increasingly being used in bioinformatics (for reviews see~cite{Almeida2013,Zou2013, Qiu2010,Taylor2010}). Specifically, for sequence analysis tools for read mapping~\cite{Schatz2009}, duplicate removal~\cite{Jourdren2012}, and variant calling~\cite{Langmead2009, McKenna2010} were released over recent years.

@@

GWAS\cite{Huang2013, Guo2014}

\url{https://mahout.apache.org/}

@@

\section*{Methods}
\subsection*{Datasets}
Overview see Table~\ref{datasets}

>>>>

As a result of these big data challenges, MapReduce approaches are increasingly being used in bioinformatics (for reviews see~cite{Almeida2013,Zou2013, Qiu2010,Taylor2010}). This is especially the case for sequence analysis tasks, such as read mapping~\cite{Schatz2009}, duplicate removal~\cite{Jourdren2012}, and variant calling~\cite{Langmead2009, McKenna2010} or Genome Wide Analysis Study based tasks~\cite{Huang2013, Guo2014}.

At the same time, sophisticated machine learning methodologies are increasingly adapted to utilise MapReduce paradigms overcoming callings that arise due to their iterative nature~\cite{Chu2009}. Specifically, the Mahout project\url{https://mahout.apache.org/} has been developed extensively~\cite{Ranger2007, Owen2011} and was successfully applied in the clinical informatics space~\cite{Dong2013}.

We therefore link the two areas by providing an interface between Mahout and the standard variant data format, VCF~\cite{1KG2012}, which opens up the application of Mahout's different machine learning algorithms to be applied to genotype-based tasks.   
To demonstrate the capability we cluster variant datasets from the 1000 genomes project to determine population structure as well as ancestry using different algorithms available in Mahout. In the first two sections we benchmark the performance and accuracy of Mahout's implementations against standard R based implementations on a reduced version of the data and in the last section we investigate the performance on the full dataset, demonstrating how seamless Mahout can transit to population-scale tasks.

@@

@@

\section*{Methods}
\subsection*{Datasets}
Trio data from the 1000 genomes project -> determine in the clustering can identify ancestry 
Sequence data from hapmap samples -> determine ethnic groups
Scaled down version of the hap map data (Chr1) for R to run on
Overview see Table~\ref{datasets}
